---
title: "BIOS7345 Lab 3"
Subtitle: "Centering"
author: "Sarah Lotspeich"
date: "27 September 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Frank's regression package
library(rms)

#all hail pipes
library(magrittr)
library(dplyr)
```

With the `Cereal.csv` data (on the GitHub), 

```{r}
dat <- read.csv("https://raw.githubusercontent.com/sarahlotspeich/Fall2019_BIOS7345_Labs/master/Data/Cereal.csv", header=TRUE, stringsAsFactors=FALSE)
```

use linear regression to regress cereal rating on calories and sugar without an interaction (use an intercept). You may use either `lm()` or `rms::ols()`. 

```{r}
mod <- 
```

Give the interpretation of the parameter estimates. Try to use the `mod` object created above to dynamically code your coefficient estimates instead. 

```{r}

```

  - $\beta_0:$ 
  - $\beta_1:$ 
  - $\beta_2:$ 

*Pop quiz:* what do you call it when you estimate based on values that are far outside the observed (or realistic) range for your predictors? 
  
Does this intercept interpretation make sense? 

Now, let's try centering both `sugars` and `calories` at their mean values

```{r}
dat$sugars_centered <- 
dat$calories_centered <- 
```

and refit the model above for `rating ~ calories_centered + sugars_centered`.

```{r}
mod_cent <- 
```

Given an interpretation of the intercept estimate. 

  - $\beta_0$: 

After centering, we should have that $\hat{\beta}_0 = \bar{y}$. Check that this is true.

```{r}

```

How do the effects of `sugars` and `calories` change with centering? 

  - $\beta_1:$ 
  - $\beta_2:$ 

```{r}

```

What about the standard errors of these effects? How can we get the standard errors directly from the model object (e.g. not printing it or just typing `mod`)? Hint: the `vcov()` function is key.

```{r}

```

Round the covariance matrix of the centered model. What do we notice about the some of the off-diagonal entries?

```{r}

```

What is happening here? Think about the [1,2] element of the covariance matrix.

\begin{eqnarray*}
Cov(\hat{\beta}_0, \hat{\beta}_1) &=& ?
\end{eqnarray*}

Now, divide the centered predictors by their standard deviations (to standardize them)

```{r}
dat$sugars_standard <- 
dat$calories_standard <-
```

and refit the model above for `rating ~ calories_standard + sugars_standard`.

```{r}
mod_standard <- 
```

Given an interpretation of the intercept estimate. 

  - $\beta_0$: 
  
What do we notice about the intercept for the standardized model? 
  
Compare the $R^2$ values for the original, centered, and standardized models. What do you see? Hint: you can access the $R^2$ of a model fit using the `rms` package using `$stats`. 

```{r}

```

Show how to obtain the coefficients from the centered/scaled model as a function of the centered/unscaled coefficients and the standard deviation of the predictor variables (e.g. obtain $\pmb{\beta}_{cent,scaled}$ from $\pmb{\beta}_{cent,unscaled}$ and $s^2_{xi}$). 

```{r}

```

Do the same for the SEs. 

```{r}

```
