---
title: "BIOS7345 Lab # 6"
subtitle: "Sums of squares"
author: "Sarah Lotspeich"
date: "18 Oct 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Suppose researchers conduct a study measuring the concentration of abnormal toxins in a specific kind of human tissue. They are interested in knowing whether differences exist based on gender or race. Data are taken from 90 individuals, n which each observation includes a response (continuous $y_i$), gender (male/ female), and race (White/ Black/ Other). 

# Balanced design
Use the following `R` code to simulate data using a balanced design: 

```{r}
n <- 90
gender <- as.factor(rep(c("male","female"), each = n/2))
race <- as.factor(rep(rep(c("W", "B", "O"), each = n/6),2))
mu <- 50 + 0.5*(gender == "male") + 0.5*(race == "B") + 0*(race == "O") + 1*(gender =="male")*(race == "O")
set.seed(1)
y <- rnorm(n, mean = mu, sd = 2)
```

## Model 1(a)
Calculate the estimated regression coefficients and corresponding p-values for the model `y \sim race + gender + race*gender` using the `lm()` function. 

```{r}

```

## Model 1(b)
Calculate the estimated regression cofficients and corresponding p-values for the model  `y \sim gender + race + gender*race` (i.e. switch the order of the main effects) using the `lm()` function.

```{r}

```

How do the estimates and p-values for Models 1(a) and 1(b) compare? 

```{r}

```

Calculate the Type I SS and corresponding p-values for Models 1(a) and 1(b). 

```{r}

```

How do they compare? 

```{r}

```

For Model 1(a), calculate the Type II and Type III sums of squares and corresponding p-values using the `car` package. Pay attention to the `type = ` option in the `Anova()` function. 

```{r}

```

How do they compare to the Type I SS?

```{r}

```

Is gender associated with the concentration of toxins? 

```{r}

```

# Unbalanced design
Use the following `R` code to simulate data using an unbalanced design: 

```{r}
n <- 90
gender <- as.factor(rep(c("male","female"), each = n/2))
race <- as.factor(c(rep("W",10),rep("B",20),rep("O",15),rep("W",5),rep("B",10),rep("O",30) ))
mu <- 50 + 0.5*(gender == "male") + 0.5*(race == "B") + 0*(race == "O") + 1*(gender =="male")*(race == "O")
set.seed(345)
y <- rnorm(n, mean = mu, sd = 2)
```

## Model 2(a)
Calculate the estimated regression coefficients and corresponding p-values for the model `y \sim race + gender + race*gender` using the `lm()` function. 

```{r}

```

## Model 2(b)
Calculate the estimated regression cofficients and corresponding p-values for the model  `y \sim gender + race + gender*race` (i.e. switch the order of the main effects) using the `lm()` function.

```{r}

```

How do the estimates and p-values for Models 2(a) and 2(b) compare? 

```{r}

```

Calculate the Type I SS and corresponding p-values for Models 2(a) and 2(b). 

```{r}

```

How do they compare? 

```{r}

```

For Model 2(a), calculate the Type II and Type III sums of squares and corresponding p-values using the `car` package. Pay attention to the `type = ` option in the `Anova()` function. 

```{r}

```

How do they compare to the Type I SS?

```{r}

```

Is gender associated with the concentration of toxins? 

```{r}

```

# Test for association with gender
Specify the appropriate $\pmb{C}$ matrix to test whether gender is associated with the concentration of toxins using $H_0: \pmb{C\beta} = \pmb{0}$. This is the strategy used in `rms::anova()` with the `test = 'Chisq'` option. 

```{r}

```